|

### 2023-10-27 chapter_2/src/roberts_empirical_chapter_1.qmd
grant application stuff

---
title: |
  Does Color Convey Political Information?
short-title: Color as Political Information
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/book_project>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
  Do the colors red and blue convey political information? While decisions of political branding are assumed to be significant for electoral outcomes, we have few theories about the cognitive processes underlying the processing of such information and whether these decisions influence realized political behavior. Building upon interdisciplinary perspectives, this chapter introduces the snap-judgment model of visual political information processing. The theory argues that color allows individuals to form pre-cognitive attitudes of a political object and that these pre-cognitive attitudes shape latter and more cognitively laborious information processing and behavior. Meaning, that we can activate partisan attachments automatically through information as simple as color. In the first study, I talk to a political marketing firm and use a pixel-by-pixel classification of images from over 1000 yard signs for elections in the U.S. House of Representatives. Evidence suggests that in districts where Democrats have an electoral advantage, there is more blue on the yard signs in the district. To examine the cognitive processes, I use a novel experimental design by tracking cursor movements and limiting participant's ability to view a yard sign to only 5000ms. I find that participants do make these associations when features of the yard sign are held constant and find that Republicans prefer candidates with red yard signs and that Democrats prefer candidates with blue yard signs.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - visual information
  - information processing
  - partisanship
date: today
bibliography: "../../assets/candidate_support_chapter_references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

@@

Do colors provide political information that shape attitudes among the mass public? There are two goals that I am hoping to achieve in this chapter. The first is to convince the reader that the colors red and blue convey affiliative information. The second is that these connections between colors and political groups can shape political outcomes.

Many understand that colors are influential to shaping attitudes and behaviors in a number of ways. Color theorists and scholars of marketing understand color as having important implications for reactions to a product and resulting consumer behavior. For example, @kuo_impact_2023 demonstrate that hotel rooms that use more "cool colors" (blues, greens, etc.) evoke positive affect for potential customers and that these positive affective responses are predictive of their eventual choice to book the room. Colors mean a varying number of things, however. As a gross, yet intelligible example, yellow sweaters relative to yellow bodily fluids can elicit strong and very different responses.

The role of visuals in politics is receiving more attention among interdisciplinary scholars in fields like communication, political science, and the psychological sciences. While many of these studies examine fascinating but complicated forms of visual information such as symbology [@williams_whats_2022] and visual framing [@grabe_image_2009], we have yet to integrate these interdisciplinary perspectives with existing models of political information processing to understand how these visual forms of information act as cues [see @lilleker_power_2019] but also what the consequences are for political behavior among the mass public.

Building upon descriptive evidence suggesting that Democratic campaigns use the color blue more than Republican campaigns [@williams_whats_2022] and that Republican voters express a preference for red during election season [@schloss_politics_2014], I present a snap-judgment model of politically-relevant visual information processing that outlines the cognitive processes that drive these associations and how those associations drive important political outcomes. I use theories of associative memory and spreading activation [see @collins_spreading-activation_1975] to conceptualize the association of the colors red and blue with Republicans and Democrats as two interconnected nodes. Informed by conceptualizations of these connections as carrying affective information informed by perspectives on biological survival reaching back to Darwin [@ralph_neuroscience_2018], I argue that when these nodes are activated by exposure to these colors in political contexts, other related nodes become active and contain affective information. As partisanship evokes strong affective reactions for many members of the public [@iyengar_affect_2012], this activation elicits an affective reaction. Considering that most visual information and that these processes of association occur pre-cognitively [@ames_impression_2012;@fazio_attitudes_2007;@sander_cambridge_2013], I argue that the elicited affective response to such information can be "transferred" [see @morris_activation_2003;@taber_illusion_2016] to later post-cognitive processing of more traditional forms of political information such as stated policy positions. The implication being that these associations can inform reactions to political information and thus political behavior pre-cognitively.
<!--
This chapter applies the snap-judgment model to a simple case: the use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [@bucy_editors_2021]. Without such foundations, we would be jumping into the study of complex cases and circumstances without a firm grasp of how things matter on a more fundamental level.

Before getting into things, it is worth noting that while there indeed is empirical and experiential evidence suggesting that Republicans use the color red and Democrats use the color blue [@schloss_politics_2014;@williams_whats_2022], there is currently no theoretical framework that explains how these connections happen at the individual level. As a side note, one might find it unsatisfying that I do not address *how* these associations between color and the parties came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election (with news networks consistently using those colors in campaign coverage [@elving_color_2014]), it is not clear if this was a deliberate choice, or whether this was yet another symptom of an increasingly polarized public. An undertaking to address the specifics of this part of American political history would likely be a different book project.
-->

To preview, in this chapter I propose two studies to evaluate the claims I make about yard signs. Yard signs are a simple yet effective source of campaign branding for political candidates as they convey information that is often useful for voters' partisanship [@campbell_american_1969]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a design and substantive standpoint. From a design standpoint, as they are static forms of party branding they have fewer moving parts, which helps isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#study-1), I find descriptive evidence suggesting that campaigns use the colors red and blue on their yard signs at different rates depending on the success of Republicans and Democrats according to previous election returns. It is unclear, however, whether these patterns are based on elite perceptions of how to communicate partisanship to voters or whether campaigns have successfully identified a cognitive process linking the colors red and blue with Republicans and Democrats. Accordingly, in [Study 2](#study-2), I conduct an online experiment that examines whether the public notice color and whether their attitudes are shaped by the presence of "Republican red" and "Democratic blue" on the yard signs of a political candidate. I find that these associations do exist among the public. Specifically, individuals consistently and strongly associate Republicans with the color red and the Democrats with the color Blue. As a result, partisan individuals' willingness to vote for a candidate can be dramatically different depending on whether the color used in the yard sign aligns with their partisan identification.

While these two studies highlight endogenous processes between campaigns and voters, to be clear, I am not arguing that either of these patterns only one way. In fact, I believe that there is a necessary feedback loop between the parties and their bases of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_voter-centric_2020], and the parties are motivated to distinguish themselves [@lee_insecure_2016;@clifford_how_2020] as a result of their high levels of ideological and affective polarization [@dietrich_using_2021;@enders_issues_2021]. I leave teasing out directionality for subsequent analyses.

These findings bring much needed evidence in support of the conventional wisdom that the colors red and blue convey political information. The implication of the theory and these findings are that partisan attachments are automatically activated through things as seemingly simple and ubiquitous as color. These activated partisan attachments then have causal effects for attitude expression and political behavior. Given the existing literature demonstrating low-levels of political engagement among the mass public [see @hibbing_stealth_2002] as well as the current hyper-polarized era [see @iyengar_affect_2012], this chapter's argument and findings suggests that voters need not consume any substantive political information before expressing a willingness to cast a vote for a candidate.

@@

The rise of television consumption changed the focus on particular mediums for political communication scholars [@hall_jamieson_creating_2014]. In a similar way, the rise of image-based social media has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_editors_2021]. However, today news organizations and politicians are responding to the ubiquitous use of image-dominant social media platforms like TikTok and Instagram, by joining and posting on such platforms. Scholars need to make this transition as well, integrating the role of simple visual information into theories of political information (abbreviated as *pip*) and attitude formation.

@@

What remains unclear is whether these patterns in the United States are meaningful. While political psychologists are increasingly interdisciplinary and rely on insights from fields like neuroscience, it remains relatively unclear whether and how color can convey complex information about politics. Further, it remains unclear whether such a simple type of information has enough efficacy to shape political outcomes such as vote intention. This chapter is a first cut at addressing these open questions.

>>>>

---
title: |
  Does Color Convey Political Information?
short-title: Color as Political Information
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/book_project>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
  We are used to seeing red and blue everywhere in campaign seasons, but do these colors actually convey political information? While decisions of political branding are assumed to be significant for electoral outcomes, we have few theories about the cognitive processes underlying the processing of such information and whether these decisions influence realized political behavior. Building upon interdisciplinary perspectives, in this chapter I first introduce the snap-judgment model of visual political information processing. The theory argues that color allows individuals to form pre-cognitive attitudes of a political object, and that these pre-cognitive attitudes shape latter and more cognitively laborious information processing and behavior -- that is, that individuals can activate partisan attachments automatically through information as simple as color. I then present two studies in dialogue with this theory, both of which are inspired by conversations with political marketing professionals. Study one focuses on what we might expect neighborhoods and communities to look like, consistent with my theory; I use a pixel-by-pixel classification of images from over 1000 yard signs for elections in the U.S. House of Representatives. My findings suggest that in districts where Democrats have an electoral advantage, that there is more blue on the yard signs in the district. Study two takes aim at cognitive processes via a controlled experimental setting. I use a novel design, tracking cursor movements and limiting participants' abilities to view a treatment image to only 5000ms. I find that participants make color associations when features of a yard sign are held constant, and that Republicans prefer candidates with red yard signs, and Democrats prefer candidates with blue yard signs.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - visual information
  - information processing
  - partisanship
date: today
bibliography: "../../assets/candidate_support_chapter_references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

@@

Do colors actually shape political attitudes among the mass public? There are two goals that I am hoping to achieve in this chapter. The first is to convince the reader that the colors red and blue convey consistent, affiliative information. The second is that these connections between colors and political groups can indeed affect political outcomes.

Many understand -- in a general sense -- that "colors matter" when it comes to attitudes and behaviors. Color theorists and scholars of marketing understand color as having important implications for reactions to a product and resulting consumer behavior. For example, @kuo_impact_2023 demonstrate that hotel rooms that use more "cool colors" (blues, greens, etc.) evoke positive affect for potential customers and that these positive affective responses are predictive of their eventual choice to book the room. Colors mean a varying number of things, however.

The role of visuals in politics is receiving more attention among interdisciplinary scholars in fields like communication, political science, and the psychological sciences. While many of these studies examine fascinating but complicated forms of visual information such as symbology [@williams_whats_2022] and visual framing [@grabe_image_2009], we have yet to integrate these interdisciplinary perspectives with existing models of political information processing -- this will enable us to better understand how these visual forms of information act as cues [see @lilleker_power_2019], and what their potential consequences might be for political behavior among the mass public.

In this chapter, I build upon descriptive evidence suggesting that Democratic campaigns use the color blue more than Republican campaigns [@williams_whats_2022] and that Republican voters express a preference for red during election season [@schloss_politics_2014]. I present a snap-judgment model of politically-relevant visual information processing that outlines a) the cognitive processes that drive these associations and b) notes how those associations drive important political outcomes. I draw upon theories of associative memory and spreading activation [see @collins_spreading-activation_1975] to conceptualize the association of the colors red and blue with Republicans and Democrats, thinking of them as as two interconnected nodes. I conceptualize these connections as carrying affective information -- an idea informed by perspectives on biological survival reaching back to Darwin [@ralph_neuroscience_2018]. When these nodes are activated by exposure to these colors in political contexts, other related nodes become active and contain affective information. As partisanship evokes strong affective reactions for many members of the public [@iyengar_affect_2012], this activation elicits an affective reaction. Considering that most visual information -- including these processes of association -- occur pre-cognitively [@ames_impression_2012;@fazio_attitudes_2007;@sander_cambridge_2013], I argue that the elicited affective response to such information can be "transferred" [see @morris_activation_2003;@taber_illusion_2016] to later post-cognitive processing of more traditional forms of political information such as stated policy positions. The implication of this is that these associations can *pre-cognitively* inform reactions to political information and political behavior

To preview, in this chapter I propose two studies to evaluate the claims I make about yard signs. Yard signs are a simple yet effective source of campaign branding for political candidates, as they convey information that is often useful for voters' partisanship [@campbell_american_1969]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a design and substantive standpoint. From a design standpoint, as a static form of party branding they have fewer moving parts, which helps isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#study-1), I find descriptive evidence suggesting that campaigns use the colors red and blue on their yard signs -- in districts -- at different rates depending on the success of Republicans and Democrats in previous elections. This paints a picture of the outside world that is consistent with my theory. However, this study does not make clear whether these patterns are based on elite perceptions of how to communicate partisanship to voters, or whether campaigns have successfully identified a cognitive process linking the colors red and blue with Republicans and Democrats. Accordingly, in [Study 2](#study-2), I conduct an online experiment that examines whether individuals notice color, and whether their attitudes are shaped by the presence of "Republican red" and "Democratic blue" on the yard signs of a political candidate. I find that these associations do exist among respondents. Specifically, individuals consistently and strongly associate Republicans with the color red and the Democrats with the color blue. I present experimental evidence that partisan individuals' willingness to vote for a candidate may be different depending on whether the color used in a yard sign aligns with their partisan identification.

While these two studies highlight endogenous processes between campaigns and voters, to be clear, I wish to be clear that I am not arguing that either of these patterns works in one direction. Rather, in all likelihood there is a necessary feedback loop between the parties and their bases of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_voter-centric_2020], and the parties are motivated to distinguish themselves [@lee_insecure_2016;@clifford_how_2020] as a result of their high levels of ideological and affective polarization [@dietrich_using_2021;@enders_issues_2021].

Still, together, the findings bring much needed evidence in support of the conventional wisdom that the colors red and blue convey political information. The implication of the theory and these findings are that partisan attachments are automatically activated through things as seemingly simple and ubiquitous as color. These activated partisan attachments then have causal effects for attitude expression and political behavior. Given the existing literature demonstrating low-levels of political engagement among the mass public [see @hibbing_stealth_2002], as well as the hyper-polarized nature of contemporary politics [see @iyengar_affect_2012], this chapter's argument and findings reinforce the idea that voters need not consume any substantive political information before expressing a willingness to cast a vote for a candidate.

@@

The rise of television consumption changed the focus on particular mediums for political communication scholars [@hall_jamieson_creating_2014]. In a similar way, the rise of image-based social media has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_editors_2021]. However, today news organizations and politicians are responding to the ubiquitous use of image-dominant social media platforms like TikTok and Instagram by joining and posting on such platforms. Scholars need to make this transition as well, integrating the role of simple visual information into theories of political information (abbreviated as *pip*) and attitude formation.

@@

What remains unclear is whether these patterns in the United States are meaningful. While political psychologists are increasingly interdisciplinary and rely on insights from fields like neuroscience, we have little systematic evidence about whether and how color can convey complex information about politics. Further, it remains to be seen whether such a simple type of information has enough potency to shape political outcomes such as vote intention. This chapter is a first cut at addressing these open and important questions.
